These lecture notes grew out of class notes provided for the students
in CSCI 104 (``Data Structures and Object-Oriented Design'') 
at the University of Southern California in Fall of 2013.
The class is typically taken in the second semester of freshman year
or the first semester of sophomore year. Students are expected to be
familiar with structural programming in the C++ programming language,
the basics of dynamic memory management, and recursion.
Brief refreshers on these two specific topics are included at the
beginning of the notes, and typically covered for about one lecture
each, but a student not already familiar with these concepts is likely
to struggle in the class.

These notes represent the specific way in which we like to present the
material, by motivating object-oriented design primarily from the
point of view of implementing data structures. There is perhaps
somewhat more focus on analysis and allusions to advanced topics than
in a typical programming-heavy data structures course for undergraduates.
The notes are, at least at present, not intended to replace an actual
detailed textbook on data structures. 
We currently use the book ``Data Abstraction and Problem Solving'' by
Carrano and Henry.

The notes are based on lecture notes taken by the CSCI 104 course
producers in Fall 2013:
Chandler Baker, Douglass Chen, Nakul Joshi, April Luo, Peter Zhang,
Jennie Zhou.
Douglass was the main notetaker, in charge of about half of the notes
himself, and coordinating the others' notes as well.
The notes were subsequently expanded significantly by David Kempe,
and then expanded further and reorganized again for subsequent semesters,
when they took essentially their current form.